A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-23 text/introduction.tex
Revise

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with special constructs for
multi-stage programming, including brackets and espace, which are
(hygienic) quasi-quotation, and \texttt{run}, which is similar to
\texttt{eval} in Lisp, and cross-stage persistence
(CSP)~\cite{MetaML}.  Programmers can easily write code generators by
using these features.  Moreover, MetaOCaml is equipped with a powerful
type system for safe code generation and execution.  The notion of
code types is introduced to prevent code values that represent
ill-typed expressions from being generated.  For example, a quotation
of expression \texttt{1 + 1} is given type \texttt{int code}; a
quotation of \texttt{abs} is given type \texttt{(int -> int) code};
and a code-generating function, which takes a code value \(c\) as an
argument and returns \(c \texttt{ + } c\), is given type \texttt{int
  code -> int code} so that it cannot be applied to, say, a quotation
of \texttt{"Hello"}, which is given type \texttt{string
  code}.  Ensuring safety for \verb|run| is more challenging because
basic code types do not guarantee that execution of code values never
results in unbound variable errors.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem and developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values with a type soundness theorem (for a pure
subset of MetaOCaml).

@@

However, the type system, which is based on the Hindley--Milner
polymorphism~\cite{Milner78JCSS}, is not strong enough to guarantee
invariants beyond simple types.  For example, Kiselyov~\cite{8384206}
demonstrates specialization of vector/matrix computation with respect
to the sizes of vectors and matrices in MetaOCaml but the type system
cannot prevent such specialized functions from being applied to
vectors and matrices of different sizes.

@@

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specialization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes an integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector |$n$\verb| -> Vector |$n$\verb|) code|.

@@

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, another calculus,
which formalizes the design of MetaOCaml more faithfully.  Our
technical contribution are summarized as follows:
\begin{itemize}
\item We give the formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

@@

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.  Section
\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.  For reference, we give the full definition of \LMD
\AI{and proofs of properties}.

>>>>

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with special constructs for
multi-stage programming, including brackets and espace, which are
(hygienic) quasi-quotation, and \texttt{run}, which is similar to
\texttt{eval} in Lisp, and cross-stage persistence
(CSP)~\cite{MetaML}.  Programmers can easily write code generators by
using these features.  Moreover, MetaOCaml is equipped with a powerful
type system for safe code generation and execution.  The notion of
code types is introduced to prevent code values that represent
ill-typed expressions from being generated.  For example, a quotation
of expression \texttt{1 + 1} is given type \texttt{int code}
and a code-generating function, which takes a code value \(c\) as an
argument and returns \(c \texttt{ + } c\), is given type \texttt{int
  code -> int code} so that it cannot be applied to, say, a quotation
of \texttt{"Hello"}, which is given type \texttt{string
  code}.  Ensuring safety for \verb|run| is more challenging because
code types per by themselves do not guarantee that the execution of code values never
results in unbound variable errors.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem, developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values, and proved a type soundness theorem (for a formal calculus \(\lambda^\alpha\) modeling a pure subset of MetaOCaml).

@@

However, the type system, which is based on the Hindley--Milner
polymorphism~\cite{Milner78JCSS}, is not strong enough to guarantee
invariants beyond simple types.  For example, Kiselyov~\cite{8384206}
demonstrates specialization of vector/matrix computation with respect
to the sizes of vectors and matrices in MetaOCaml but the type system
of MetaOCaml cannot prevent such specialized functions from being applied to
vectors and matrices of different sizes.

@@

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specialization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes a (nonnegative) integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector |$n$\verb| -> Vector |$n$\verb|) code|.

@@

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, which formalizes
the design of MetaOCaml more faithfully.  Dependent types are based on
\LLF~\cite{attapl}, which has one of the simplest form of dependent
types.  Our technical contribution are summarized as follows:
\begin{itemize}
\item We give a formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

@@

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.
Section~\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.  For reference, we give the full definition of \LMD \AI{and
  proofs of properties} in Appendix.

### 2019-06-23 text/properties.tex

Substitution Lemma in \LMD{} is little more complicated than usual
because there are eight judgment forms and two kinds of substitution.
Term Substitution Lemma states that term substitution $[z \mapsto M]$
preserves derivability of judgments and Stage Substitution Lemma
states similarly for stage substitution $[\alpha\mapsto A]$.

We let $\mathcal{J}$ stand for judgments $K \iskind @A$, $\tau::K@A$,
$M:\tau@A$, $K \E J @ A$, $\tau \E \sigma @ A$, and
$M \E N : \tau @ A$.  Substitutions $\mathcal{J}[z \mapsto M]$ and
$\mathcal{J}[\alpha \mapsto A]$ are defined in a straightforward
manner.  Using these notations, the two substitution lemmas are stated as follows:

\begin{lemma}[Term Substitution]
  If $\G, z:\xi@B, \D \V \mathcal{J}$ and $\G\V P:\xi @B$, then $\G, \D[z \mapsto P] \V \mathcal{J}[z \mapsto P]$.  Similarly, if $\V \G, z:\xi@B, \D$ and
  $\G\V P:\xi @B$, then $\V \G, \D[z \mapsto P]$.
\end{lemma}

@@

\begin{proof}
  Each item is strengthen by statements about type equivalence.
  For example, the first statement is augmented by
  \begin{quotation}
    If $\G \V \rho \E (\Pi x:\sigma.\tau) : K @A$, then there exist
    $\sigma'$ and $\tau'$ such that $\rho = \Pi x:\sigma'.\tau'$ and
    $\G \V \sigma \E \sigma' : K @A$ and
    $\G, x:\sigma@A\V \tau \E \tau' : J @A$.
  \end{quotation}
  and its symmetric version.  Then, they are proved simultaneously by induction on derivations.

  Similarly for the others.
 \qed
\end{proof}

@@

Thanks to Substitution Lemmas and Inversion Lemmas, we can prove Preservation easily.

@@

\begin{proof}
	First, there are three cases for $M \longrightarrow M'$.
	They are $M \longrightarrow_\beta M'$, $M \longrightarrow_\Lambda M'$, and $M \longrightarrow_\blacklozenge M'$.
	For each case, we can use straightforward induction on typing derivations.
	Difficult cases are \TApp, \TTBL, and \TIns.
	We need Inversion Lemmas for them.
\end{proof}

@@

Strong Normalization is also an important property, which guarantees that
no typed term has an infinite reduction sequence.
We prove this theorem by translating \LMD to the simply typed lambda calculus~\cite{harper1993framework}.

@@

Confluence is a property that any reduction sequences from one typed term converge.
Since we have proved Strong Normalization, we can use Newman's Lemma~\cite{} to prove Confluence.

@@

\begin{proof}
  We can prove by straightforward induction on typing derivations.
  Main cases are \TApp, \TTBL, and \TIns, where we use Inversion.
  \qed
\end{proof}

>>>>

Substitution Lemma in \LMD{} is a little more complicated than usual
because there are eight judgment forms and two kinds of substitution.
Term Substitution Lemma states that term substitution $[z \mapsto M]$
preserves derivability of judgments and Stage Substitution Lemma
states similarly for stage substitution $[\alpha\mapsto A]$.

We let $\mathcal{J}$ stand for the judgments $K \iskind @A$, $\tau::K@A$,
$M:\tau@A$, $K \E J @ A$, $\tau \E \sigma @ A$, and
$M \E N : \tau @ A$.  Substitutions $\mathcal{J}[z \mapsto M]$ and
$\mathcal{J}[\alpha \mapsto A]$ are defined in a straightforward
manner.  Using these notations, the two substitution lemmas are stated as follows:

\begin{lemma}[Term Substitution]
  If $\G, z:\xi@B, \D \V \mathcal{J}$ and $\G\V P:\xi @B$, then $\G, (\D[z \mapsto P]) \V \mathcal{J}[z \mapsto P]$.  Similarly, if $\V \G, z:\xi@B, \D$ and
  $\G\V P:\xi @B$, then $\V \G, (\D[z \mapsto P])$.
\end{lemma}

@@

\begin{proof}
  Each item is strengthen by statements about type equivalence.
  For example, the first statement is augmented by
  \begin{quotation}
    If $\G \V \rho \E (\Pi x:\sigma.\tau) : K @A$, then there exist
    $\sigma'$ and $\tau'$ such that $\rho = \Pi x:\sigma'.\tau'$ and
    $\G \V \sigma \E \sigma' : K @A$ and
    $\G, x:\sigma@A\V \tau \E \tau' : J @A$.
  \end{quotation}
  and its symmetric version.  Then, they are proved simultaneously by induction on derivations.
  Similarly for the others.
 \qed
\end{proof}

@@

Thanks to Term/Stage Substitution and Inversion, we can prove Preservation easily.

@@

\begin{proof}
	First, there are three cases for $M \longrightarrow M'$.
	They are $M \longrightarrow_\beta M'$, $M \longrightarrow_\Lambda M'$, and $M \longrightarrow_\blacklozenge M'$.
	For each case, we can use straightforward induction on typing derivations.
	% Difficult cases are \TApp, \TTBL, and \TIns.
	% We need Inversion Lemmas for them.
\end{proof}

@@

Strong Normalization is also an important property, which guarantees that
no typed term has an infinite reduction sequence.
Following standard proofs (see, e.g., \cite{harper1993framework}), we prove this theorem by translating \LMD to the simply typed lambda calculus.

@@

Confluence is a property that any reduction sequences from one typed term converge.
Since we have proved Strong Normalization, we can use Newman's Lemma~\cite{DBLP:books/daglib/0092409} to prove Confluence.

@@

\begin{proof}
  We can prove by straightforward induction on typing derivations.
%  Main cases are \TApp, \TTBL, and \TIns, where we use Inversion.
  \qed
\end{proof}

### 2019-06-23 text/related-work.tex

MetaOCaml is a programming language with quoting, unquoting, run, and
CSP.  Kiselyov give many applications of MetaOCaml in \cite{8384206},
which includes filtering in signal processing, matrix-vector product,
and DSL compiler.

@@

Theoretical studies on multi-stage programming owe a lot to seminal
work by Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, which found Curry-Howard
correspondence between multi-stage calculi and modal logic.  In
particular, Davies' $\lambda^\circ$~\cite{davies1996temporal} has been
a basis for multi-stage calculi with quasi-quotation.  $\lambda^\circ$
did not have operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP}
enhanced and improved $\lambda^\circ$ towards the development of a
type-safe multi-stage calculus with quasi-quotation, run, and CSP,
which were proposed by Taha and Sheard as constructs for multi-stage
programming~\cite{MetaML}.
% Benaissa et al.~\cite{benaissa1999logical} study the relationship between a multi-stage calculus and category theory or modal logic.
% Taha and Sheard introduced run and CSP to a multi-stage calculus in 
Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a
type sound manner and formed a basis of earlier versions of MetaOCaml.
Different approaches to type-safe multi-stage programming with
slightly different constructs for composing and running code values have been studied by Kim, Yi,
and Calcagno~\cite{DBLP:conf/popl/KimYC06} and Nanevski and Pfenning~\cite{DBLP:journals/jfp/NanevskiP05}.

Later, Tsukada and Igarashi~\cite{Tsukada} found correspondence
between a variant of \(\lambda^\alpha\) called $\lambda^\TW$
and modal logic and showed that run could be represented as a special
case of application of a transition abstraction ($\Lambda\alpha$) to
the empty sequence $\varepsilon$.  Hanada and
Igarashi~\cite{Hanada2014} developed \LTP as an extension
$\lambda^\TW$ with CSP.
% and discuss code residualization which
% allows us to dump the quoted code into an external file.

@@

There is much work on dependent types and most of them are affected by
the pioneering work by Martin-L\"{o}f~\cite{martin1973intuitionistic}.
Among many dependent type systems such as
$\lambda^\Pi$~\cite{Meyer1986}, Calculus of
Constructions~\cite{coquand:inria-00076024}, and Edinburgh
LF~\cite{harper1993framework}, we base our work on \LLF~\cite{attapl}
(which is quite close to $\lambda^\Pi$ and Edinburgh LF) due to its
simplicity.  It has been well known that dependent types are useful to
express detailed properties of data structure at the type level such
as the size of data structures~\cite{Xi98} and typed abstract syntax
trees~\cite{DBLP:conf/dsl/LeijenM99,DBLP:conf/popl/XiCC03}.  The
vector addition discussed in Section~\ref{sec:formal} is also such an
example.

>>>>

MetaOCaml is a programming language with quoting, unquoting, run, and
CSP.  Kiselyov~\cite{8384206} describes many applications of MetaOCaml, including filtering in signal processing, matrix-vector product,
and DSL compiler.

@@

Theoretical studies on multi-stage programming owe a lot to seminal
work by Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, who found Curry-Howard
correspondence between multi-stage calculi and modal logic.  In
particular, Davies' $\lambda^\circ$~\cite{davies1996temporal} has been
a basis for several multi-stage calculi with quasi-quotation.  $\lambda^\circ$
did not have operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP}
enhanced and improved $\lambda^\circ$ towards the development of a
type-safe multi-stage calculus with quasi-quotation, run, and CSP,
which were proposed by Taha and Sheard as constructs for multi-stage
programming~\cite{MetaML}.
% Benaissa et al.~\cite{benaissa1999logical} study the relationship between a multi-stage calculus and category theory or modal logic.
% Taha and Sheard introduced run and CSP to a multi-stage calculus in 
Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a
type sound manner and formed a basis of earlier versions of MetaOCaml.
Different approaches to type-safe multi-stage programming with
slightly different constructs for composing and running code values have been studied by Kim, Yi,
and Calcagno~\cite{DBLP:conf/popl/KimYC06} and Nanevski and Pfenning~\cite{DBLP:journals/jfp/NanevskiP05}.

Later, Tsukada and Igarashi~\cite{Tsukada} found correspondence
between a variant of \(\lambda^\alpha\) called $\lambda^\TW$
and modal logic and showed that run could be represented as a special
case of application of a transition abstraction ($\Lambda\alpha.M$) to
the empty sequence $\varepsilon$.  Hanada and
Igarashi~\cite{Hanada2014} developed \LTP as an extension
$\lambda^\TW$ with CSP.
% and discuss code residualization which
% allows us to dump the quoted code into an external file.

@@

There is much work on dependent types and most of them are affected by
the pioneering work by Martin-L\"{o}f~\cite{martin1973intuitionistic}.
Among many dependent type systems such as
$\lambda^\Pi$~\cite{Meyer1986}, The Calculus of
Constructions~\cite{coquand:inria-00076024}, and Edinburgh
LF~\cite{harper1993framework}, we base our work on \LLF~\cite{attapl}
(which is quite close to $\lambda^\Pi$ and Edinburgh LF) due to its
simplicity.  It has been well known that dependent types are useful to
express detailed properties of data structure at the type level such
as the size of data structures~\cite{Xi98} and typed abstract syntax
trees~\cite{DBLP:conf/dsl/LeijenM99,DBLP:conf/popl/XiCC03}.  The
vector addition discussed in Section~\ref{sec:formal} is also such an
example.
